---
title: "STAT-427_FinalProject_Kimb"
author: "J.Kimbrough"
date: "2025-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pressure, echo=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(dplyr)
library(dbplyr)
library(RSQLite)
library(ggthemes)
library(ggplot2)
library(class)
library(caret)
library(glmnet)
library(ISLR2)
library(MASS)
library(boot)
library(tree)
library(randomForest)
library(car)
```

```{r}
directory_path <- "C:/Users/jkimb/OneDrive/Desktop/School/Classes/VII. Spring 2025 (LAST SEM BABY)/ONL_STAT-427_StatisticalMachineLearning/Projects"

setwd(directory_path)
```

```{r}
mta_data_raw <- read_csv("./data/MTA_Daily_Ridership_Data__2020_-_2025.csv", na = "missing")

view(mta_data_raw)
glimpse(mta_data_raw)
```


# ---- Cleaning Data -----

```{r}
anyNA(mta_data_raw)
any(sapply(mta_data_raw, function(x) any(x=="")))
any(sapply(mta_data_raw, function(x) any(x == -999)))
```

```{r}
df <- mta_data_raw %>%
  mutate(Date = mdy(Date)) %>%
  mutate(across(where(is.character), ~ na_if(., "missing"))) %>%
  mutate(across(where(is.numeric), ~ na_if(., -999))) %>%
  mutate(across(where(is.character), ~ na_if(., ""))) %>% 
  na.omit()

head(df)
```


# --- Exploratory Data Analysis ---

```{r}
colnames(df)
```

```{r}
hist(df$`Subways: Total Estimated Ridership`, main = "Total Estimated Subway Ridership")
```

```{r}
hist(df$`Buses: Total Estimated Ridership`, main = "Total Estimated Buses Ridership")
```

```{r}
hist(df$`LIRR: Total Estimated Ridership`, main = "LIRR Total Estimated Ridership")
```


```{r}
hist(df$`Metro-North: Total Estimated Ridership`, main = "Metro-North Total Estimated Ridership")
```


```{r}
hist(df$`Staten Island Railway: Total Estimated Ridership`, main = "SIR Total Estimated Ridership")
```

### Looks Skewed Left

```{r}
plot(df$`Subways: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`, main = "Subway % of Comparable Pre-Pandemic Day v Subway Total Estimated Ridership", xlab = "Subway % of Comparable Pre-Pandemic Day", ylab = "Subway Total Estimated Ridership")
```

###

```{r}
plot(df$`Buses: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "Buses Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "Buses Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```

###

```{r}
plot(df$`LIRR: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "LIRR Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "LIRR Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```

###


```{r}
plot(df$`Metro-North: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "Metro-North Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "Metro-North Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```

###


```{r}
plot(df$`Access-A-Ride: Total Scheduled Trips`, df$`Subways: Total Estimated Ridership`, main = "Access-A-Ride Total Scheduled Trips v Subway Total Estimated Ridership", xlab = "Access-A-Ride Total Scheduled Trips", ylab = "Subway Total Estimated Ridership")
```

###


```{r}
plot(df$`Bridges and Tunnels: Total Traffic`, df$`Subways: Total Estimated Ridership`, main = "Bridges and Tunnels Total Traffic v Subway Total Estimated Ridership", xlab = "Bridges and Tunnels Total Traffic", ylab = "Subway Total Estimated Ridership")
```

###


```{r}
plot(df$`Staten Island Railway: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "SIR Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "SIR Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```

###


```{r}
cor(df$`Subways: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### Moderate positive correlation. As subway ridership returns to pre-pandimc levels, total subway ridership tends to increase. 


```{r}
cor(df$`Buses: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

### Strong postive correlation. Higher bus ridership seems strongly associated with higher subway ridership. 


```{r}
cor(df$`Buses: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### Modeate positive correlation. When bus ridership returns to pre-pandemic levels, subway ridership also tends to be higher. 


```{r}
cor(df$`LIRR: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

### A very strong positive correlation. LIRR ridership and subway ridership seem strongly associated with each other. 


```{r}
cor(df$`LIRR: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### A weak, but positive correlation. As LIRR ridership returns to pre-pandemic levels, it doesn't appear to affect Subway total estimated ridership significantly. 


```{r}
cor(df$`Metro-North: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

### A strong positive correlation. Metro-north ridership and subway ridership seem strongly associated with each other. 

```{r}
cor(df$`Metro-North: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### Moderate positive correlation. As Metro-north ridership returns to pre-pandemic levels, it somewhat tracks with subway usage.

```{r}
cor(df$`Access-A-Ride: Total Scheduled Trips`, df$`Subways: Total Estimated Ridership`)
```

### A strong positive correlation. Access-A-Ride trips and subway ridership seem strongly associated with each other. 


```{r}
cor(df$`Access-A-Ride: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### Moderately positive correlation. As Access-A-Ride usage returns to pre-pandemic levels, total subway ridership tends to increase.


```{r}
cor(df$`Bridges and Tunnels: Total Traffic`, df$`Subways: Total Estimated Ridership`)
```

### Moderately positive correlation. Higher vehicle traffic seems associated with higher subway ridership.


```{r}
cor(df$`Bridges and Tunnels: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### Moderately positive correlation. As Bridge and Tunnel traffic returns to pre-pandemic levels, total subway ridership seems to somewhat increase.

```{r}
cor(df$`Staten Island Railway: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

### A strong positive correlation. Staten Island Railway and Subway ridership seem strongly associated with each other. 


```{r}
cor(df$`Staten Island Railway: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

### A weak, but positive correlation. As SIR ridership returns to pre-pandemic levels, it doesn't appear to affect Subway total estimated ridership significantly. 



# Regression Question

## Can we predict total estimated subway ridership based on pandemic/post-pandemic trends and while considering external factors such as economic indicators?

## --- Multiple Linear Regression ---

```{r}
# Subways: Total Estimated Ridership

full_model <- lm(`Subways: Total Estimated Ridership` ~ ., data = df)

summary(full_model)

```

### The model explains about 99.45% of the variability in subway ridership, which is extremely high. With an F-stat of 22,890 and a p-value < 2.2e-16, the overall model is highly statistically significant.



```{r}
plot(full_model)
```


```{r}
#NO Date, Bridges and Tunnels: % of Comparable Pre-Pandemic Day
reduced_model <- lm(`Subways: Total Estimated Ridership` ~ . -Date -`Bridges and Tunnels: % of Comparable Pre-Pandemic Day`, data = df)
summary(reduced_model)
```

### This model still explains about 99.45% of the variability in subway ridership, which is extremely high. The F-stat is still pretty high at 26,630 and a low p-value. The overall model is still statistically significant. 


```{r}
plot(reduced_model)
```

```{r}
step(reduced_model)
```

### No variables were dropped because AIC was already the lowest.


```{r}
anova(full_model, reduced_model)
```
### Reduced model actually worsened statistically. So, keeping full model performs better. 


```{r}
vif_full <- vif(full_model)
vif_full
```

### Some multicollinearity was detected, with `LIRR Total Estimated Ridership` having the highest vif. 



```{r}
predicted_full <- predict(full_model, df)

rmse_full <- sqrt(mean((df$`Subways: Total Estimated Ridership` - predicted_full)^2))
rmse_full
```

```{r}
mse_full <- (mean((df$`Subways: Total Estimated Ridership` - predicted_full)^2))
mse_full
```

### The RMSE shows that there is an average daily prediction error of approximately 78,900 riders.

```{r}
set.seed(1234)
train_control <- trainControl(method = "cv", number = 10)

model_cv <- train(
  `Subways: Total Estimated Ridership` ~ .,
  data = df,
  method = "lm",
  trControl = train_control
)

print(model_cv)
```

### Similar to predicted RMSE for the full model, the cross-validated RMSE shows that there is an average daily prediction error of approximately 80,000 riders. The R^2 is very high, and suggests that the model explains over 99% of the variability in subway ridership. 

```{r}
options(scipen = 999)
actual <- df$`Subways: Total Estimated Ridership`
predicted <- predicted_full

plot(actual, predicted,
     xlab = "Actual Subway Ridership (Billions)",
     ylab = "Predicted Subway Ridership (Billions)",
     main = "Predicted vs. Actual Subway Ridership",
     pch = 19, col = "steelblue")
abline(a = 0, b = 1, col = "red", lwd = 2)
```


```{r}
response_var <- "Subways: Total Estimated Ridership"
predictor_vars <- c("Buses: Total Estimated Ridership", 
                    "LIRR: Total Estimated Ridership", 
                    "Metro-North: Total Estimated Ridership",
                    "Access-A-Ride: Total Scheduled Trips", 
                    "Bridges and Tunnels: Total Traffic", 
                    "Staten Island Railway: Total Estimated Ridership")

df_clean <- na.omit(df[, c(response_var, predictor_vars)])


x <- as.matrix(df_clean[, predictor_vars])
y <- df_clean[[response_var]]


set.seed(168)
train_indices <- sample(1:nrow(x), 0.8 * nrow(x))
x_train <- x[train_indices, ]
y_train <- y[train_indices]
x_test  <- x[-train_indices, ]
y_test  <- y[-train_indices]


ridge_model <- cv.glmnet(x_train, y_train, alpha = 0)
ridge_pred <- predict(ridge_model, s = ridge_model$lambda.min, newx = x_test)


lasso_model <- cv.glmnet(x_train, y_train, alpha = 1)
lasso_pred <- predict(lasso_model, s = lasso_model$lambda.min, newx = x_test)


full_model <- lm(`Subways: Total Estimated Ridership` ~ ., data = df)
predicted_full <- predict(full_model, df)


mlr_plot_df <- data.frame(Actual = df$`Subways: Total Estimated Ridership`, 
                          Predicted = predicted_full, 
                          Model = "Linear Regression")

ridge_plot_df <- data.frame(Actual = y_test, 
                            Predicted = as.numeric(ridge_pred), 
                            Model = "Ridge Regression")

lasso_plot_df <- data.frame(Actual = y_test, 
                            Predicted = as.numeric(lasso_pred), 
                            Model = "Lasso Regression")


combined_df <- rbind(mlr_plot_df, ridge_plot_df, lasso_plot_df)


ggplot(combined_df, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +
  labs(
    title = "\n         Predicted vs Actual Subway Ridership",
    x = "Actual Ridership (Billions)",
    y = "Predicted Ridership (Billions)",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```




# Classification Question

## Can we categorize daily ridership levels as “low,” “average,” or “high” based on total ridership numbers?


```{r}
### combining ridership 
df$total_ridership <- (df$`Subways: Total Estimated Ridership`+
                         df$`Buses: Total Estimated Ridership`+
                         df$`LIRR: Total Estimated Ridership`+
                         df$`Metro-North: Total Estimated Ridership`+
                         df$`Staten Island Railway: Total Estimated Ridership`)
```



```{r}

quantiles <- quantile(df$total_ridership, probs = c(0.33, 0.66))

df$total_ridership_level <- cut(
  df$total_ridership,
  breaks = c(-Inf, quantiles[1], quantiles[2], Inf),
  labels = c("Low", "Average", "High")
)
```



```{r}
predictors <- c(
  "Subways: % of Comparable Pre-Pandemic Day",
  "Buses: % of Comparable Pre-Pandemic Day",
  "LIRR: % of Comparable Pre-Pandemic Day",
  "Metro-North: % of Comparable Pre-Pandemic Day",
  "Access-A-Ride: % of Comparable Pre-Pandemic Day",
  "Bridges and Tunnels: % of Comparable Pre-Pandemic Day",
  "Staten Island Railway: % of Comparable Pre-Pandemic Day"
)
```



## ---- KNN ---- 
```{r}
df_knn <- na.omit(df[, c(predictors, "total_ridership_level")])

```


```{r}
set.seed(1234)

Z = sample(nrow(df_knn), .5*nrow(df_knn))

train = df_knn[Z, predictors]

test = df_knn[-Z, predictors]

cl = df_knn$total_ridership_level[Z]
test_cl = df_knn$total_ridership_level[-Z]

```

```{r}
Yhat <- knn(train, test, cl, k = 3)
```

```{r}
conf_matx <- table(Predicted = Yhat, Actual = test_cl)
conf_matx 
```

```{r}
accuracy <- sum(diag(conf_matx)) / sum(conf_matx)
accuracy
```

### Errors mostly happen around the Low v Average and Average v High boundary. Pretty high accuracy. 


```{r}
class_rates <- numeric(100)

for(k in 1:100){
  Yhat_k <- knn(train, test, cl, k = k)
  
  conf_matx_k <- table(Predicted = Yhat_k, Actual = test_cl)
  
  class_rates[k] <- sum(diag(conf_matx_k)) / sum(conf_matx_k)
}
```

```{r}
best_k <- which.max(class_rates)
best_k
```


```{r}
plot(1:100, class_rates, type = "b", pch = 19, col = "steelblue", main = "Best K",
     xlab = "K Value", ylab = "Classification Accuracy")

abline(v = best_k, col = "red", lty = 7)
```

### k = 1 performed the best but this could be indicative that the model may be overfitting. 




























































