---
title: "STAT-427_FinalProject_Kimb"
author: "J.Kimbrough"
date: "2025-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pressure, echo=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(dplyr)
library(dbplyr)
library(RSQLite)
library(ggthemes)
library(ggplot2)
library(class)
library(caret)
library(glmnet)
library(ISLR2)
library(MASS)
library(boot)
library(tree)
library(randomForest)
```

```{r}
directory_path <- "C:/Users/jkimb/OneDrive/Desktop/School/Classes/VII. Spring 2025 (LAST SEM BABY)/ONL_STAT-427_StatisticalMachineLearning/Projects"

setwd(directory_path)
```

```{r}
mta_data_raw <- read_csv("./data/MTA_Daily_Ridership_Data__2020_-_2025.csv", na = "missing")

view(mta_data_raw)
glimpse(mta_data_raw)
```


# ---- Cleaning Data -----

```{r}
anyNA(mta_data_raw)
any(sapply(mta_data_raw, function(x) any(x=="")))
any(sapply(mta_data_raw, function(x) any(x == -999)))
```

```{r}
df <- mta_data_raw %>%
  mutate(Date = mdy(Date)) %>%
  mutate(across(where(is.character), ~ na_if(., "missing"))) %>%
  mutate(across(where(is.numeric), ~ na_if(., -999))) %>%
  mutate(across(where(is.character), ~ na_if(., ""))) %>% 
  na.omit()

head(df)
```


# --- Exploratory Data Analysis ---

```{r}
colnames(df)
```

```{r}
hist(df$`Subways: Total Estimated Ridership`, main = "Total Estimated Subway Ridership")
```

```{r}
hist(df$`Buses: Total Estimated Ridership`, main = "Total Estimated Buses Ridership")
```

```{r}
hist(df$`LIRR: Total Estimated Ridership`, main = "LIRR Total Estimated Ridership")
```


```{r}
hist(df$`Metro-North: Total Estimated Ridership`, main = "Metro-North Total Estimated Ridership")
```


```{r}
hist(df$`Staten Island Railway: Total Estimated Ridership`, main = "SIR Total Estimated Ridership")
```


```{r}
plot(df$`Subways: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`, main = "Subway % of Comparable Pre-Pandemic Day v Subway Total Estimated Ridership", xlab = "Subway % of Comparable Pre-Pandemic Day", ylab = "Subway Total Estimated Ridership")
```


```{r}
plot(df$`Buses: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "Buses Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "Buses Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```


```{r}
plot(df$`LIRR: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "LIRR Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "LIRR Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```


```{r}
plot(df$`Metro-North: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "Metro-North Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "Metro-North Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```


```{r}
plot(df$`Access-A-Ride: Total Scheduled Trips`, df$`Subways: Total Estimated Ridership`, main = "Access-A-Ride Total Scheduled Trips v Subway Total Estimated Ridership", xlab = "Access-A-Ride Total Scheduled Trips", ylab = "Subway Total Estimated Ridership")
```

```{r}
plot(df$`Bridges and Tunnels: Total Traffic`, df$`Subways: Total Estimated Ridership`, main = "Bridges and Tunnels Total Traffic v Subway Total Estimated Ridership", xlab = "Bridges and Tunnels Total Traffic", ylab = "Subway Total Estimated Ridership")
```

```{r}
plot(df$`Staten Island Railway: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`, main = "SIR Total Estimated Ridership v Subway Total Estimated Ridership", xlab = "SIR Total Estimated Ridership", ylab = "Subway Total Estimated Ridership")
```


```{r}
cor(df$`Subways: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Subways: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Buses: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Buses: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`LIRR: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`LIRR: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Metro-North: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Metro-North: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Access-A-Ride: Total Scheduled Trips`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Access-A-Ride: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Bridges and Tunnels: Total Traffic`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Bridges and Tunnels: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Staten Island Railway: Total Estimated Ridership`, df$`Subways: Total Estimated Ridership`)
```

```{r}
cor(df$`Staten Island Railway: % of Comparable Pre-Pandemic Day`, df$`Subways: Total Estimated Ridership`)
```

# Regression Question

## Can we predict total estimated subway ridership based on pandemic/post-pandemic trends and while considering external factors such as economic indicators?

## --- Multiple Linear Regression ---

```{r}
# Subways: Total Estimated Ridership

full_model <- lm(`Subways: Total Estimated Ridership` ~ ., data = df)

summary(full_model)

```

```{r}
plot(full_model)
```


```{r}
#NO Date, Bridges and Tunnels: % of Comparable Pre-Pandemic Day
reduced_model <- lm(`Subways: Total Estimated Ridership` ~ . -Date -`Bridges and Tunnels: % of Comparable Pre-Pandemic Day`, data = df)
summary(reduced_model)
```

```{r}
plot(reduced_model)
```


```{r}
anova(full_model, reduced_model)
```

```{r}
library(car)
vif_full <- vif(full_model)
vif_full
```

```{r}
vif_reduced <- vif(reduced_model)
vif_reduced
```



```{r}
predicted_full <- predict(full_model, df)

rmse_full <- sqrt(mean((df$`Subways: Total Estimated Ridership` - predicted_full)^2))
rmse_full
```

```{r}
predicted_reduced <- predict(reduced_model, df)

rmse_reduced <- sqrt(mean((df$`Subways: Total Estimated Ridership` - predicted_reduced)^2))
rmse_reduced
```


```{r}
set.seed(1234)
library(boot)

cv_error_full <- cv.glm(df, full_model, K = 10)
cv_error_full$delta[1]  
sqrt(cv_error_full$delta[1])  
```

```{r}
set.seed(1234)
library(boot)

cv_error_reduced <- cv.glm(df, reduced_model, K = 10)
cv_error_reduced$delta[1]  
sqrt(cv_error_reduced$delta[1])
```


# Classification Question

## Can we categorize daily ridership levels as “low,” “average,” or “high” based on total ridership numbers?


```{r}
### combining ridership 
df$total_ridership <- (df$`Subways: Total Estimated Ridership`+
                         df$`Buses: Total Estimated Ridership`+
                         df$`LIRR: Total Estimated Ridership`+
                         df$`Metro-North: Total Estimated Ridership`+
                         df$`Staten Island Railway: Total Estimated Ridership`)
```

```{r}
mean_total <- mean(df$total_ridership)
sd_total <- sd(df$total_ridership)
```

```{r}
df$total_ridership_level <- cut(
  df$total_ridership,
  breaks = c(-Inf, mean_total - sd_total, mean_total + sd_total, Inf),
  labels = c("Low", "Average", "High")
)
```



```{r}
predictors <- c(
  "Subways: % of Comparable Pre-Pandemic Day",
  "Buses: % of Comparable Pre-Pandemic Day",
  "LIRR: % of Comparable Pre-Pandemic Day",
  "Metro-North: % of Comparable Pre-Pandemic Day",
  "Access-A-Ride: % of Comparable Pre-Pandemic Day",
  "Bridges and Tunnels: % of Comparable Pre-Pandemic Day",
  "Staten Island Railway: % of Comparable Pre-Pandemic Day"
)
```



## ---- KNN ---- 
```{r}
df_knn <- na.omit(df[, c(predictors, "total_ridership_level")])

```



```{r}
set.seed(1234)

Z = sample(nrow(df_knn), .5*nrow(df_knn))

train = df_knn[Z, predictors]

test = df_knn[-Z, predictors]

cl = df_knn$total_ridership_level[Z]
test_cl = df_knn$total_ridership_level[-Z]

```

```{r}
Yhat <- knn(train, test, cl, k = 3)
```

```{r}
conf_matx <- table(Predicted = Yhat, Actual = test_cl)
conf_matx 
```

```{r}
accuracy <- sum(diag(conf_matx)) / sum(conf_matx)
accuracy
```


```{r}
class_rates <- numeric(100)

for(k in 1:100){
  Yhat_k <- knn(train, test, cl, k = k)
  
  conf_matx_k <- table(Predicted = Yhat_k, Actual = test_cl)
  
  class_rates[k] <- sum(diag(conf_matx_k)) / sum(conf_matx_k)
}
```

```{r}
best_k <- which.max(class_rates)
best_k
```

```{r}
plot(1:100, class_rates, type = "b", pch = 19, col = "blue",
     xlab = "K Value", ylab = "Classification Accuracy")

abline(v = best_k, col = "red", lty = 3)
```






























































